\section{Gr\o nlund and Pettie Applied to 3POL}

In Paper~\ref{paper:3pol-algorithm}
we focus on the computational complexity of 3POL\@. Since 3POL contains 3SUM,
an interesting question is whether a generalization of Gr\o nlund and Pettie's
3SUM algorithm exists for 3POL\@. If this is true, then we might wonder whether
we can ``beat'' the $O(n^{11/6}) = O(n^{1.833\ldots})$ combinatorial bound of Raz,
Sharir and de Zeeuw~\cite{RSZ15} with nonuniform algorithms
(see \S\ref{sec:history:pol:combinatorics}).
%
We give a positive
answer to both questions: we design
a uniform
$O(n^2 {(\log \log n)}^{3/2} / {(\log n)}^{1/2})$-time
algorithm
and
a nonuniform
$O(n^{12/7+\epsilon}) = O(n^{1.7143})$-time algorithm
for 3POL\@.
%
To prove our uniform result, we present a fast algorithm for the Polynomial
Dominance Reporting (PDR) problem, a far reaching generalization of the
Offline Dominance Reporting problem.
%
As the algorithm for Offline Dominance Reporting and its
analysis by Chan~\cite{Cha08} is used in fast algorithms for all-pairs shortest
paths, (min,+)-convolutions, and 3SUM, we expect this new algorithm will have
more applications.

To make the exposition of our results simpler,
we study two different variants of the 3POL problem.
The first variant is the 3POL problem as defined in the previous chapter
(\S\ref{sec:history:pol:3pol}).
We recall its definition here.
\ProblemPOLImplicit*
The second variant is a special case of the 3POL problem where we
restrict the trivariate polynomial $F$ to have the form $F(a,b,c) = f(a,b) -
c$. We call it the explicit 3POL problem because the dependency on the third
variable is explicitly given:
%
\input{text/definition/3pol-explicit}

Similarly to Gr\o nlund and Pettie~\cite{GP18}, we consider both nonuniform
and uniform models of computation.
%
For the nonuniform model, Gr\o nlund and Pettie consider linear
decision trees.
%
Because linear decision trees are not powerful enough to even solve the
problems we are looking at,
in this contribution, we consider
\emph{bounded-degree algebraic decision trees (ADT)},
an algebraic generalization of linear decision trees
(see~\S\ref{sec:models-of-computation:algorithms:trees}).
%
For the uniform model we consider the real-RAM model with only the four
arithmetic operators (see \S\ref{sec:models-of-computation:algorithms:ram}),
the same model as in~\cite{GP18}.

In \S\ref{sec:algo:explicit:nonuniform} and \S\ref{sec:algo:explicit:uniform},
we first design uniform and nonuniform algorithms for explicit 3POL.
%
\input{text/theorem/3pol-algorithm/nonuniform-explicit}
\input{text/theorem/3pol-algorithm/uniform-explicit}

In \S\ref{sec:algo:implicit:nonuniform} and \S\ref{sec:algo:implicit:uniform},
we show that those algorithms can be adapted to solve the more general 3POL
problem.
%
\input{text/theorem/3pol-algorithm/nonuniform-implicit}
\input{text/theorem/3pol-algorithm/uniform-implicit}

Table~\ref{tor:3pol-algorithm} gives a summary of our results.

\begin{table}
\centering
\caption{Algorithmic results for 3POL and its explicit variant.}
\label{tor:3pol-algorithm}
\begin{tabular}{|c|c|c|}
	\hline

	Reference & Model & Complexity \\
	\hline
	\hline
	\S\ref{sec:algo:explicit:nonuniform} (\ref{thm:explicit:nonuniform}),
	\S\ref{sec:algo:implicit:nonuniform} (\ref{thm:implicit:nonuniform}) &
	ADT &
	\(O(n^{12/7 + \epsilon})\)
	\\

	\hline

	\S\ref{sec:algo:explicit:uniform}(\ref{thm:explicit:uniform}),
	\S\ref{sec:algo:implicit:uniform}(\ref{thm:implicit:uniform}) &
	real-RAM &
	$O(n^2 {(\log \log n)}^{3/2} / {(\log n)}^{1/2})$
	\\

	\hline

\end{tabular}
\end{table}

\paragraph{Applications}

Our results can be applied to many algebraic degeneracy testing problems, such
as GPT, a well known 3SUM-hard problem for which no subquadratic algorithm is
known (see \S\ref{sec:history:pol}).
%
Raz, Sharir
and de Zeeuw results on the 3POL problem~\cite{RSZ15} can be applied to obtain
a combinatorial bound of $O(n^{11/6})$ on the
number of collinear triples when the input points are known to be lying on
a constant number of polynomial curves, provided those curves are neither
lines nor cubic curves.
%
A corollary of our first result is that
GPT where the input points are constrained to lie on
$o({(\log n)}^{1/6}/{(\log \log n)}^{1/2})$
constant-degree polynomial curves (including lines and cubic curves)
admits a subquadratic real-RAM algorithm and
a strongly subquadratic bounded-degree algebraic decision tree.
Interestingly, both reductions from 3SUM to GPT on 3 lines (map $a$ to $(a,0)$,
$b$ to $(b,2)$, and $c$ to $(-\frac c2, 1)$) and from 3SUM to GPT on a
cubic curve (map $a$ to $(a^3,a)$, $b$ to $(b^3,b)$, and $c$ to $(c^3,c)$)
construct such special instances of GPT (see
\S\ref{sec:history:pol:reductions-from-ksum} for details on those reductions).
This constitutes the first step towards closing the major open question of
whether GPT can be solved in subquadratic time.
%
To further convince the reader of the expressive power of the 3POL problem,
we also give reductions
from the problem of counting triples of points spanning
unit circles (\S\ref{sec:paper:3pol-algorithm:application:circles}),
from the problem of counting triples of points spanning unit area
triangles (\S\ref{sec:paper:3pol-algorithm:application:triangles}),
and
from the problem of counting collinear triples in any dimension
(\S\ref{sec:paper:3pol-algorithm:application:gpt}).


\paragraph{A Remark}

The algorithms we present manipulate polynomial expressions.
%
In computational geometry, it is customary to assume the real-RAM model can be
extended to allow the computation of roots of constant degree polynomials.
We distance ourselves from this practice and take particular care
of using the real-RAM model and the bounded-degree algebraic decision tree
model with only the four arithmetic operators (see Chapter~\ref{chapter:etr} on
the Existential Theory of the Reals).
