\section{Better Encodings for 3SUM}

\todo{Add definition of the 3SUM type function}

%In the 3SUM problem, we are given an array of numbers as input and are asked
%whether any three of them sum to \(0\). In the mid-nineties, this problem was
%identified as a bottleneck of many important problems in geometry, such as
%detection of affine degeneracies or motion planning~\cite{GO95}. Since then, it
%has become a central problem in fine-grained complexity theory~\cite{PW10}. It
%has long been conjectured to require $\Omega(n^2)$ time. In 2014, it was shown
%to be solvable in $o(n^2)$ time, but no algorithm with running time
%$O(n^{2-\delta})$ with constant $\delta>0$ is known~\cite{GP18}.

%Lower bounds exist in restricted models of computation. Most notably,
%$\Omega(n^2)$ \(3\)-linear queries are needed to solve 3SUM~\cite{Er99a}, and
%nontrivial lower bounds have also been proven for slightly more powerful linear
%decision trees~\cite{AC05}. However, in a recent breakthrough contribution,
%Kane, Lovett, and Moran showed that 3SUM could be solved using $O(n\log^2 n)$
%6-linear queries~\cite{KLM18}, hence within a $O(\log n)$ factor of the
%information-theoretic lower bound.

%Linear decision trees are examples of {\em nonuniform algorithms}, in which we
%are allowed to have different algorithms for different input sizes.
%Algebraic decision trees generalize linear decision trees
%by allowing decision based on the sign of constant-degree polynomials at each
%node~\cite{SY82}.

%Any decision tree identifying the 3SUM type of a 3SUM instance yields a concise
%encoding of this 3SUM type:
%just write down the outcome of the successive tests. Knowing the decision tree
%by convention, this sequence of bits is sufficient to recover the sign of any
%triple.

%The question we consider here is how to make such a representation efficient,
%in the sense that not only does it use merely a few bits, but the answer to any
%triple query can be recovered efficiently. Understanding the interplay between
%nonuniform algorithms and such data structures hopefully sheds light on the
%intrinsic structure of the problem.

In Paper~\ref{paper:3sum-encoding},
we consider the following problem: given three sets of real numbers, output a
word-RAM data structure from which we can efficiently recover the sign of the
sum of any triple of numbers, one in each set.

This problem is similar to the problem of encoding the order
type of a finite set of points studied in Paper~\ref{paper:order-type-encoding}.
While this contribution showed that it was
possible to achieve slightly subquadratic space and logarithmic query time, we
show here that we can encode the \emph{3SUM type} of \(n\) numbers using
\(\tilde{O}(n^{3/2})\) bits to allow for constant time queries in the
word-RAM.
%
We also study lower and upper bounds of other natural 3SUM type encodings.

As there are only $O(n^3)$ queries, a table
of size $n^3 \log_2 3 + O(1)$ bits suffices to give constant query time~\cite{DPT10}.
%
This can be improved to $O(n^2\log n)$ bits of space by
storing for each pair $(i,j)$ the values
\(k_<(i,j) = \max \{ 0\}\cup \{k \colon\, a_i + b_j + c_k < 0\}\) and
\(k_>(i,j) = \min \{ n+1\}\cup \{k \colon\, a_i + b_j + c_k > 0\}\).
For a query \((i,j,k)\), we compare \(k\) against the values \(k_<(i,j)\) and \(k_>(i,j)\)
to recover \(\chi(i,j,k)\) in \(O(1)\) time. All \(k_<(i,j)\) and \(k_>(i,j)\)
can be computed in \(O(n^2)\) time via the classic quadratic time algorithm for
3SUM.
%: if \(k \leq k_<(i,j)\), then \(\chi(i,j,k)=-1\), if \( k_<(i,j)< k < k_>(i,j)\),
%then \(\chi(i,j,k)=0\), and if \( k_>(i,j)\leq k\), then \(\chi(i,j,k)=+1\).
%The representation takes \( O(n^2 \log n \) bits, and each query can be answered by comparing
%pairs of indices, which takes \(O(1)\) time.

One seemingly simple representation is to store the numbers in $A$, $B$ and
$C$; however these are reals and thus we need to make them representable using
a finite number of bits.
In \S\ref{s:numbers} we show that a minimal integer representation of a
3SUM instance may require $\Theta(n)$ bits per value, which would give
rise to a $O(n)$ query time and $O(n^2)$ space, which is far from
impressive.
%
\input{text/theorem/3sum-encoding/bitsize}

In Paper~\ref{paper:order-type-encoding}
we studied
the problem of given a set of $n$ lines, to create an
encoding of them so that the orientation of any triple (the \emph{order type})
can be determined; our problem is a special case of this where the
lines only have three slopes.
Can we do better for the case of 3SUM? We answer this in the affirmative.

Again, for ease of presentation, we state our results in terms of the encoding
definition of \S\ref{sec:models-of-computation:data-structures:encodings} that
we restate here.
%
\DefinitionEncoding*

In \S\ref{s:space} we show how to use an optimal $O(n \log n)$ bits of
space with a polynomial query time.
\input{text/theorem/3sum-encoding/space-optimal}

Finally, in \S\ref{s:sscqt} we show
how to use $\tilde{O}(n^{3/2})$ space to achieve $O(1)$-time queries.
\input{text/theorem/3sum-encoding/gp}

Table~\ref{tor:3sum-encoding} gives a summary of past and new results.

\begin{table}
\centering
\caption{Past results and new contributions on 3SUM type encoding.}%
\label{tor:3sum-encoding}
\begin{tabular}{|c|c|c|c|}
\hline
Reference & Query time & Space (bits) & Preprocessing \\ \hline \hline
Trivial & $O(1)$ & $O(n^3)$ & $O(n^3)$ \\ \hline
Almost trivial & $O(1)$ & $O(n^2 \log n)$ & $O(n^2)$ \\ \hline
Order type (Paper~\ref{paper:order-type-encoding}) & $O(\log n)$ & $O(\frac{n^2 \log^2 \log n}{\log n})$ & $O(n^2) $\\ \hline
Order type (Paper~\ref{paper:order-type-encoding}) & $O(\frac{\log n}{\log \log n})$ & $O(\frac{n^2 }{\log^{1-\epsilon} n})$ & $O(n^2)$ \\ \hline
\hline
Numerical \S\ref{s:numbers}(\ref{lem:bitsize}) & $O(n)$ & $O(n^2)$ & $n^{O(1)}$\\ \hline
Space-optimal \S\ref{s:space}(\ref{theorem:3sum-encoding:space-optimal}) & $n^{O(1)}$ & $O(n \log n)$ & $n^{O(1)}$\\ \hline
Query-optimal \S\ref{s:sscqt}(\ref{theorem:3sum-encoding:gp}) &  $O(1)$ & $\tilde{O}(n^{3/2})$ & $O(n^{2})$ \\ \hline
\end{tabular}
\end{table}
