We mainly consider the Euclidean plane \(\mathbb{R}^2\) and its
\(d\)-dimensional analog \(\mathbb{R}^d\).

\chapter{Arrangements of Hyperplanes}

An \emph{hyperplane} in \(\mathbb{R}^d\) is the set of points \(q \in \mathbb{R}^d\)
satisfying some linear equation of the form \(\sum_{i=1}^{d} p_i q_i = 0\) for
some tuple of coefficients \(p \in \mathbb{R}^d\). An \emph{affine} hyperplane has an
additional independent term \(p_0\) to get the equation \(\sum_{i=1}^{d}
p_i q_i = p_0\). In general we will drop the adjective ``affine'' and talk about
hyperplanes without specifying whether \(p_0 = 0\).

In the the plane an hyperplane is a line. In \(\mathbb{R}^3\), an hyperplane is
a plane. On the real line, an hyperplane is a point.

An hyperplane \(H\) partitions the space \(\mathbb{R}^d\) into two halfspaces,
often denoted by \(H^-\) and \(H^+\).
Naturally, \(H^-\) is defined to be the set of points such that
\(\sum_{i=1}^{d} p_i q_i \leq p_0 \) and \(H^+\) is defined to be the set of
points such that \(\sum_{i=1}^{d} p_i q_i \geq p_0 \). When we want to be more
precise as to whether the inequality is strict we can write \(H^{<0}\),
\(H^{\leq 0}\), \(H^{>0}\), or \(H^{\geq 0}\).
We also define \(H^0\) and \(H^{=0}\) as synonyms of \(H\).

A set of hyperplanes \(\mathcal{H} = \{\, H_1, H_2, \ldots, H_m\,\}\)
partitions \(\mathbb{R}^d\) into cells.
%
This partition is denoted by \(\mathcal{A}(\mathcal{H})\) and is called the
\emph{arrangement} of \(\mathcal{H}\).
%
Each sign vector \(\sigma \in \{\,
-,0,+\,\}\) corresponds to a (potentially empty) cell of this arrangement
defined as the set of points \(q\) such that \(q \in H_i^{\sigma_i}\) for all
\(H_i \in \mathcal{H}\).

\section{Counting Cells}
It is interesting to understand the behavior of such an arrangement when
\(m\) grows. The goal is to bound the number of nonempty cells an arrangement
can have. When \(m \leq d\) we can easily create patological examples where
every possible cell exists and is equal to \(3^m\). But when we fix
\(d\) and make \(m\) grow, we get another interesting behavior.

We state a theorem due to Buck~\cite{Bu43} that allows us to conclude
this chapter with the result that the decision tree complexity of sorting
\(X + Y\) is \(O(n^2)\).

The theorem goes as follows
\begin{theorem}[Buck~\cite{Bu43}]\label{theorem:xy:counting}
Consider the partition of space defined by an arrangement of $m$ hyperplanes in
$\R^d$.
The number of regions of dimension $k \le d$ is at most
\begin{displaymath}
	{m \choose d-k}
	\left(
		{m-d+k \choose 0}
		+
		{m-d+k \choose 1}
		+
		\cdots
		+
		{m-d+k \choose k}
	\right)
\end{displaymath}
and the number of regions of all dimensions is \(O(m^d)\).
\end{theorem}

To use this theorem, we model Sorting \(X + Y\) as a point location problem in an
arrangement of hyperplanes. Each instance of the problem is consisting of two
sets of real numbers $\X$ and $\Y$, both of cardinality $n$. An instance can
thus be encoded as a vertex in $\R^{2n}$. Here is a possible encoding,
\begin{displaymath}
x = (\X_1, \ldots, \X_n, \Y_1, \ldots, \Y_n).
\end{displaymath}

For every combination $(\X_{i_1}, \X_{j_1}, \Y_{i_2}, \Y_{j_2})$ of $4$
components of this vertex we want to know whether $\X_{i_1} + \Y_{i_2} \le
\X_{j_1} + \Y_{j_2}$. This can be rewritten as,
$$\X_{i_1} + \Y_{i_2} - \X_{j_1} - \Y_{j_2} \ask{\le} 0$$
and it is equivalent to asking if $x$ lies above, on or under an
hyperplane of equation $\X_{i_1} + \Y_{i_2} - \X_{j_1} - \Y_{j_2} = 0$. There
are \(O(n^4)\) of those hyperplanes, and thanks to \ref{theorem:xy:counting}
we know that an arrangement built from them would contain
\(O({(n^4)}^{2n}) = O(n^{8n})\) regions. Hence, for \(X+Y\), \(\card{\Gamma} =
O(n^{8n})\) and by using the algorithm of Fredman to find the correct
$n^2$-tuple, the decision tree complexity is
\(O(n \log n + n^2) = O(n^2)\).

In the next section we give a visual proof for a simplified version of
Buck's theorem.



\section{Pseudolines}
\dots

Pseudolines are generalized into pseudohyperplanes. We do not discuss this
generalization in this thesis.

\section{Zone Theorem}
\input{text/tool/zone-theorem}

\section{Geometric View}

Explain how any decision about polynomial predicates applied to tuples of input
real numbers is a geometry problem. Define point location.

Point location: finding the prism of the VD that contains the input point is
sufficient to answer all queries. Such a prism is defined by a linear number of
supporting queries.



\section{Point Location}
Describe the point location problem.
\input{text/definition/point-location}

\chapter{Point Configurations}

\section{Duality}
\input{text/tool/duality}

\section{Existing Encodings}
\input{text/tool/wire-diagrams}
\input{text/tool/lambda-matrices}
\input{text/tool/stars}
\input{text/tool/canonical-labelings}

\chapter{Cell Partitioning}

\section{Bottom Vertex Triangulation}
\input{text/tool/bottom-vertex-triangulation}

\section{Vertical Decomposition}
\input{text/tool/vertical-decomposition}


\chapter{Divide and Conquer}

This generalizes the idea of divide-and-conquer in sorting.

\section{\(\varepsilon\)-nets}
\input{text/tool/epsilon-nets}

\section{Cuttings}
\input{text/tool/cuttings}

\section{Hierarchical Cuttings}
\input{text/tool/hierarchical-cuttings}

\chapter{Existential Theory of the Reals}
\input{text/tool/existential-theory-of-the-reals}
