\paragraph*{\iftitlecase%
Hierarchical Cuttings\else%
Hierarchical cuttings\fi}
We encode the order type of an arrangement via
hierarchical cuttings as defined in~\cite{C93}. A cutting in \(\mathbb{R}^d\)
is a set of (possibly unbounded and/or non-full dimensional)
bounded-complexity cells that together partition \(\mathbb{R}^{d}\).
%
For our purposes, a cell is of bounded complexity if its boundary is defined by
a number of lines or pseudolines (and later, hyperplanes) of the arrangement
that depends only on the dimension, and not on the size of the arrangement.
%
A \(\frac{1}{c}\)-cutting of a set of \(n\) hyperplanes is a cutting with the
constraint that each of its cells is intersected by at most \(\frac{n}{c}\)
hyperplanes. There exist various ways of constructing \(\frac{1}{c}\)-cuttings of
size \(O(c^d)\).
\ifeurocg\else%
Those cuttings allow for efficient divide-and-conquer
solutions to many geometric problems. The hierarchical cuttings of Chazelle
have the additional property that they can be composed without multiplying the
hidden constant factors in the big-O notation. In particular, they allow
for \(O(n^d)\)-space \(O(\log n)\)-query \(d\)-dimensional point location data
structures (for constant \(d\)).
\fi%
\ifjournal
  \begin{definition}[Hierarchical Cutting]
    Given \(n\) hyperplanes in \(\mathbb{R}^d\),
    a \(\ell\)-levels hierarchical cutting of parameter \(r > 1\)
    for those hyperplanes
    is a sequence of \(\ell\) levels labeled \(0,1, \ldots, \ell - 1\)
    such that%
    \footnote{In~\cite{C93}, Chazelle refers to this parameter as
    \(r_0\) and uses \(r\) to mean \(r_0^\ell\). Here we drop the subscript for
    ease of presentation.}
    \begin{itemize}
      \item Level \(i\) has \(O(r^{2i})\) cells,
      \item Each of those cells is further partitioned into \(O(r^2)\)
        subcells,
      \item The collection of subcells is a \(\frac{1}{r^{i+1}}\)-cutting for
        the set of hyperplanes,
      \item The \(O(r^{2(i+1)})\) subcells of level \(i\) are the cells of level \(i+1\).
    \end{itemize}
  \end{definition}
  \aurelien{Use constant \(c_k\) or \(c_d\) instead of Big-Oh?}
  It is clear from reading through the various references that those
  hierarchical cuttings can be constructed for arrangements of pseudolines with
  the same properties:
  In~\cite{C93},
  Chazelle first proves a vertex-count estimation lemma
  that only relies on incidence properties of line
  arrangements~\cite[Lemma~2.1]{C93}. Then he summons a lemma from~\cite{Ma93}
  that relies on the finite VC-dimension of the range space at
  hand~\cite[Lemma 3.1]{C93}.
  Here the ground set is the set of pseudolines and the ranges are the
  subsets induced by intersections with pseudosegments.
  The VC-dimension of this range space
  is easily shown to be finite and is known to be at most
  \(8\): every arrangement of \(9\) pseudolines contains a subset of
  \(6\) pseudolines in hexagonal formation~\cite{HM94}, which cannot be
  shattered.%
  \footnote{This a quote from~\cite{BMP05}. We could not access
  the original paper.}
  %
  \aurelien{The VC-dimension is not even needed. Only enumerating ranges is
  necessary.}
  Finally, he proves a lemma for the efficient construction of
  sparse \(\varepsilon\)-nets whose correctness again only relies on incidence
  properties of line arrangements~\cite[Lemma 3.2]{C93}.
  Using those three lemmas together with bottom vertex triangulation he is
  able to prove his main result:
\else%
In the plane, hierarchical cuttings can be
constructed for arrangement of pseudolines with the same properties.
\fi

\ifjournal
\begin{lemma}[{Chazelle~\cite[Theorem 3.3]{C93}}]\label{lem:hierarchical-cutting-d}
  Given \(n\) hyperplanes in \(\mathbb{R}^d\), for any real parameter \(r >
  1\), we can construct a \(\ell\)-levels hierarchical cutting of parameter
  \(r\) for those hyperplanes in time \(O(nr^{\ell(d-1)})\).
\end{lemma}

For pseudoline arrangements, bottom vertex triangulation can be traded for
vertical decomposition.
\begin{lemma}\label{lem:hierarchical-cutting-2}
  Given \(n\) pseudolines, for any real parameter \(r > 1\), we can construct
  a \(\ell\)-levels hierarchical cutting of parameter
  \(r\) for those pseudolines in time \(O(nr^\ell)\).
\end{lemma}

In particular, we will use those lemmas with
\(\ell = \lceil \log_r \frac nt \rceil\),
for some parameter \(t\),
so that the last level of the hierarchy defines a \(\frac
tn\)-cutting whose cells are each intersected by at most \(t\) pseudolines (or
hyperplanes).

Note that in~\cite{C93} the construction is described for constant
parameter \(r\).
This restriction on the parameter is easily lifted:
We can construct a hierarchical
cutting with superconstant parameter \(r\) by constructing a hierarchical
cutting with some appropriate constant parameter \(r'\), and then skip levels that we do
not need. This is used in Section~\ref{sec:query-time} to reduce the query time
from \(O(\log n)\) to \(O(\frac{\log n}{\log \log n})\).
\fi
