\section{GPT \& 3POL}\label{sec:problem:pol}

In the plane, the General Position Testing problem (GPT) asks whether, given
\(n\) points, three of them are collinear~\cite{???}.
%
Because of some elementary algebra, the following form is equivalent:
%
\input{text/definition/gpt-2}

Note that this determinant is a degree-two polynomial in six variables
\begin{displaymath}
	x_i y_j - y_i x_j - x_i y_k + y_i x_k + x_j y_k - y_j x_k.
\end{displaymath}

In Computational Geometry it is better known as the clockwise/counterclockwise
predicate because for given points \(i\), \(j\), and \(k\) the sign of this
expression gives the orientation of the triangle those points define.
%
It is more commonly written as
\begin{displaymath}
	(x_j - x_i)(y_k - y_i) - (y_j - y_i)(x_k - x_i),
\end{displaymath}
with only five subtractions and two multiplications.

The problem generalizes to higher dimension: given \(n\) points in
\(\mathbb{R}^d\), decide whether any \(d+1\) of them lie on a common
hyperplane. Again, using algebra:
%
\input{text/definition/gpt-d}

Again, this determinant is a degree \(d\) polynomial in
\(d^2 + d\) variables.

Later in this document we mention GPT without specifying the dimension.
The problem is then assumed to be in the plane.
When we consider a different parameterization we explicitly mention it.

\subsection{\done Variants}

Similarly to the 3SUM problem, we can study a
variant of GPT where the tested triples come from different sets.
%
\input{text/definition/gpt-2-variant}

This variant is useful for a particular reduction from 3SUM.
A similar variant can be defined for the \(d\)-dimensional problem.

\subsection{\done Reductions from \(k\)-SUM}%
\label{sec:problem:pol:reductions-from-ksum}

This section is divided into two parts.
%
The first part exposes a folk reduction from \(k\)-SUM to GPT showing
GPT in \(\mathbb{R}^d\) is as hard as \((d+1\))-SUM.
%
The second part expose other interesting connections between \(k\)-SUM and GPT
that can be drawn through Vandermonde determinants, yielding equivalent
hardness proofs for GPT~\cite{Er99b}.


\subsubsection{Folklore}

Given an instance of the three-set variant of 3SUM,
we can reduce it to an instance of GPT where the input sets are on three
horizontal lines.
%
\begin{lemma}
Given three sets \(A\), \(B\), and \(C\) of \(n\) real numbers, let
\begin{align*}
    A' &= \{\, a' = (a,0) \colon\, a \in A\,\},\\
    B' &= \{\, b' = (b,2) \colon\, b \in B\,\},\\
    C' &= \{\, c' = (-c/2,1) \colon\, c \in C\,\}.
\end{align*}
Then for any \(a \in A\), \(b \in B\), and \(c \in C\), we have that
\(a+b+c=0\) if and only if \(a'\), \(b'\), and \(c'\) are collinear.
\end{lemma}
%
\begin{proof}
    The equation of a line defined by points \(a' \in A'\) and \(b' \in B'\)
    is
    \begin{align*}
	y &= \frac{2}{b-a} x - \frac{2}{b-a} a\\
	\iff x &= a + \frac{y (b-a)}{2},
    \end{align*}
    which simplifies to \(x = \frac{a+b}{2}\) when \(y = 1\).
\end{proof}

Therefore, solving GPT is at least as hard as solving 3SUM.

In the dual, this reduction is even easier to apprehend and trivially
generalizes to \(d\)-dimensional GPT.
%
\begin{lemma}
Given \(k\) sets \(S_i\) of \(n\) real numbers,
in \(\mathbb{R}^{k-1}\)
let
\(S_i'\), \(1 \leq i \leq k-1\),
be the set of \(n\) hyperplanes \(x_i = s\) for \(s \in S_i\) and let
\(S_k'\) be the set of \(n\) hyperplanes \(\sum_{i=1}^{k-1} x_i + s = 0\) for
\(s \in S_k\).
Then for any \(s_i \in S_i\), \(1 \leq i \leq k\),
we have that
\(\sum_{i=1}^{k} s_i = 0\)
the corresponding hyperplanes intersect.
\end{lemma}
%
\begin{proof}
    Trivially.
\end{proof}
%
This proves that GPT in \(\mathbb{R}^d\) is as hard as (\(d+1\))-SUM.

\subsubsection{Points on the Weird Moment Curve}

When \(d=1\), GPT is exactly the element uniqueness problem.
%
\input{text/definition/gpt-1}

One seemingly stupid way to solve this problem is to
consider the Vandermonde matrix
%
\begin{displaymath}
V(x) = \begin{pmatrix}
    1       & x_1     & x_1^2     & \dots  & x_1^{n-1}     \\
    1       & x_2     & x_2^2     & \dots  & x_2^{n-1}     \\
    \vdots  & \vdots  & \vdots    & \ddots & \vdots        \\
    1       & x_n & x_n^2 & \dots  & x_n^{n-1}
\end{pmatrix},
\end{displaymath}
%
and to compute its determinant
%
\begin{displaymath}
|V(q)| = \det\begin{pmatrix}
    1       & q_1     & q_1^2     & \dots  & q_1^{n-1}     \\
    1       & q_2     & q_2^2     & \dots  & q_2^{n-1}     \\
    \vdots  & \vdots  & \vdots    & \ddots & \vdots        \\
    1       & q_n & q_n^2 & \dots  & q_n^{n-1}
\end{pmatrix}
=
\prod_{1 \leq i < j \leq n} (q_j - q_i).
\end{displaymath}

This can be done in \(O(n \log n)\) ring operations by defining
%
\(p(x) = \prod_{i=0}^{n-1} (x-q_i)\)
%
and computing
%
\begin{align*}
{\det(V(q))}^2 &= \mathrm{disc}_x(p(x)) \\
&= {(-1)}^{\binom{n}{2}} \mathrm{res}_x(p(x), \mathrm{diff}_x(p(x))) \\
&= {(-1)}^{\binom{n}{2}} \prod_{i=0}^{n-1} \mathrm{diff}_x(p(x))(q_i),
\end{align*}
%
using fast polynomial interpolation to first find the coefficients of
\(p(x)\) and then using fast polynomial evaluation to evaluate
\(\mathrm{diff}_x(p(x))\) at all \(q_i\)~\cite{St73,Ku73a,Ku73b,ASU75}.

Consider points on the moment curve $m_d(t) = (t, t^2, \ldots, t^d)$. Given a
GPT instance consisting of points $m_d(q_i)$ on this curve the determinant
in Problem~\ref{problem:gpt-d} becomes
%
\begin{displaymath}
\det\begin{pmatrix}
1      & q_1^1 & q_1^2 & \cdots & q_1^d \\
1      & q_2^1 & q_2^2 & \cdots & q_2^d \\
\vdots & \vdots  & \vdots  & \ddots & \vdots  \\
1      & q_{d+1}^1 & q_{d+1}^2 & \cdots & q_{d+1}^d
\end{pmatrix}
=
|V(q)|.
\end{displaymath}

Consider points on the \emph{weird} moment curve $\omega_d(t) = (t, t^2, \ldots,
t^{d-1}, t^{d+1})$~\cite{Er99b}.
If our input only consists of points $\omega_d(q_i)$ on this curve then the
determinant becomes
%
\begin{displaymath}
\det\begin{pmatrix}
1      & q_0^1 & q_0^2 & \cdots & q_0^{d-1} & q_0^{d+1} \\
1      & q_1^1 & q_1^2 & \cdots & q_1^{d-1} & q_1^{d+1} \\
\vdots & \vdots  & \vdots  & \ddots & \vdots & \vdots  \\
1      & q_d^1 & q_d^2 & \cdots & q_d^{d-1} & q_d^{d+1}
\end{pmatrix} = |V(q)| \cdot \sum_{i=1}^{d+1} q_i,
\end{displaymath}
%
which is the predicate we want to test for $(d+1)$-SUM (modulo the $|V(q)|$
factor which is nonzero for distinct input numbers).

In the plane, this reduction from 3SUM to the cubic curve \(\omega_2(t) = (t,
t^3)\) is also part of folklore.
%Note that the weird moment curve in two dimensions is precisely the cubic curve
%from the reduction from 3SUM to GPT.

\subsection{Lower Bounds}

Lower bounds (Goodman Pollack Seidel? Ailon)

Ben-or applies Milnor Thom

\subsection{Algorithms}

Naive algorithm

Duality and constructing the dual arrangement

No subquadratic algorithm known, uniform or nonuniform.

\subsection{The Intermediate Problem}

We consider an algebraic generalization of the 3SUM problem: we replace the sum
function by a constant-degree polynomial in three variables $F \in
\mathbb{R}[x,y,z]$ and ask to determine whether there exists a
\emph{degenerate} triple $(a,b,c)$ of input numbers such that $F(a,b,c)=0$. We
call this new problem the \emph{3POL problem}.

\input{text/definition/3pol}

\subsection{Combinatorics}

In a series of results spanning fifteen years,
Elekes and R贸nyai~\cite{ER00},
Elekes and Szab贸~\cite{ES12},
Raz, Sharir and Solymosi~\cite{RSS14}, and
Raz, Sharir and de Zeeuw~\cite{RSZ15}
give upper bounds on the number of degenerate triples for the 3POL problem.
%
For the particular case $F(x,y,z) = f(x,y) - z$ where $f \in \mathbb{R}[x,y]$
is a constant-degree bivariate polynomial, Elekes and R贸nyai~\cite{ER00} show
that the number of degenerate triples is $o(n^2)$ unless $f$ is
\emph{special}. Special for $f$ means that $f$ has one of the two special forms
\begin{displaymath}
f(u,v)=h(\varphi(u)+\psi(v))
\qquad
\text{or}
\qquad
f(u,v)=h(\varphi(u)\cdot\psi(v)),
\end{displaymath}
where $h,\varphi,\psi$ are univariate polynomials of constant degree.
It must be noted that the 3SUM problem falls in the special category since, in
that case, \( f \) is the sum function.
%
Elekes and Szab贸~\cite{ES12} later generalized this result to a broader range
of functions $F$ using a wider definition of specialness.
%
Raz, Sharir and Solymosi~\cite{RSS14} and Raz, Sharir and de Zeeuw~\cite{RSZ15}
improved both bounds to $O(n^{11/6})$.
%
They translated the problem into an incidence problem between points and
constant-degree algebraic curves. Then, they showed that unless $f$ (or $F$) is
special, these curves have low multiplicities. Finally, they applied a theorem
due to Pach and Sharir~\cite{PS98} bounding the number of incidences between
the points and the curves.
%
\input{text/theorem/RSZ15}

Raz, Sharir and de Zeeuw~\cite{RSZ15} also look at the number of degenerate
triples for the General Position Testing problem when the input is restricted
to points lying on a constant number of constant-degree algebraic curves.
%
\input{text/theorem/RSZ15-gpt}

Nassajian Mojarrad, Pham, Valculescu and de Zeeuw~\cite{MPVd16} and
Raz, Sharir and de Zeeuw~\cite{RSZ16} proved bounds for versions of the
problem where $F$ is a $4$-variate polynomial.


\subsection{Encodings}

Pseudoline arrangements

Encodings

Canonical Labelings
