\section{Generalization of Gr\o nlund and Pettie's Results to 3POL}

We focus on the computational complexity of 3POL\@. Since 3POL contains 3SUM,
an interesting question is whether a generalization of Gr\o nlund and Pettie's
3SUM algorithm exists for 3POL\@. If this is true, then we might wonder whether
we can ``beat'' the $O(n^{11/6}) = O(n^{1.833\ldots})$ combinatorial bound of Raz,
Sharir and de Zeeuw~\cite{RSZ15} with nonuniform algorithms. We give a positive
answer to both questions: we design
a uniform
$O(n^2 {(\log \log n)}^{3/2} / {(\log n)}^{1/2})$-time
real-RAM algorithm
and
a nonuniform
$O(n^{12/7+\varepsilon}) = O(n^{1.7143})$-depth
bounded-degree algebraic decision tree
for 3POL\@.
%
To prove our uniform result, we present a fast algorithm for the Polynomial
Dominance Reporting (PDR) problem, a far reaching generalization of the
Dominance Reporting problem. As the algorithm for Dominance Reporting and its
analysis by Chan~\cite{Cha08} is used in fast algorithms for all-pairs shortest
paths, (min,+)-convolutions, and 3SUM, we expect this new algorithm will have
more applications.

Our results can be applied to many algebraic degeneracy testing problems, such
as the General Position Testing~(GPT) problem: ``Given $n$ points in the plane, do
three of them lie on a line?'' It is well known that GPT is 3SUM-hard,
and it is open whether GPT admits a subquadratic algorithm. Raz, Sharir
and de Zeeuw results on the 3POL problem~\cite{RSZ15} can be applied to obtain
a combinatorial bound of $O(n^{11/6})$ on the
number of collinear triples when the input points are known to be lying on
a constant number of polynomial curves, provided those curves are neither
lines nor cubic curves. A corollary of our first result is that
GPT where the input points are constrained to lie on
$o({(\log n)}^{1/6}/{(\log \log n)}^{1/2})$
constant-degree polynomial curves (including lines and cubic curves)
admits a subquadratic real-RAM algorithm and
a strongly subquadratic bounded-degree algebraic decision tree.
Interestingly, both reductions from 3SUM to GPT on 3 lines (map $a$ to $(a,0)$,
$b$ to $(b,2)$, and $c$ to $(\frac c2, 1)$) and from 3SUM to GPT on a
cubic curve (map $a$ to $(a^3,a)$, $b$ to $(b^3,b)$, and $c$ to $(c^3,c)$)
construct such special instances of GPT\@.
This constitutes the first step towards closing the major open question of
whether GPT can be solved in subquadratic time.
%
To further convince the reader of the expressive power of the 3POL problem,
we also give reductions from the problem of counting triples of points spanning
unit circles, from the problem of counting triples of points spanning unit area
triangles, and from the problem of counting collinear triples in any dimension.

We study two different generalizations of 3SUM\@. In the first generalization,
which we call the 3POL problem, we replace the sum function by a trivariate
polynomial of constant degree:
\ProblemPOLImplicit*
The second generalization is a special case of the 3POL problem where we
restrict the trivariate polynomial $F$ to have the form $F(a,b,c) = f(a,b) -
c$. We call it the explicit 3POL problem because the dependency on the third
variable is explicitly given:
\input{text/definition/3pol-explicit}
We will first design algorithms for this easier problem.
We design both uniform and nonuniform algorithms.

\input{text/theorem/3pol-algorithm/nonuniform-explicit}
\input{text/theorem/3pol-algorithm/uniform-explicit}

Those algorithms can then be adapted to work for the more general 3POL
problem.

\input{text/theorem/3pol-algorithm/nonuniform-implicit}
\input{text/theorem/3pol-algorithm/uniform-implicit}

\paragraph{Models of Computation}

Similarly to Gr\o nlund and Pettie~\cite{GP18}, we consider both nonuniform
and uniform models of computation.
%
For the nonuniform model, Gr\o nlund and Pettie consider linear
decision trees, where one is only allowed to manipulate the input numbers
through linear queries to an oracle. Each linear query has constant cost and
all other operations are free but cannot inspect the input.
%
In this paper, we consider
\emph{bounded-degree algebraic decision trees (ADT)}~\cite{R72,Y81,SY82},
an algebraic generalization of linear decision trees,
as the nonuniform model. In a bounded-degree algebraic decision tree, one
performs constant cost branching operations that amount to test the sign of
a constant-degree polynomial of the input numbers. Again,
operations not involving the input are free.
%
For the uniform model we consider the real-RAM model with only the four
arithmetic operators.

The algorithms we present manipulate polynomial expressions.
%
In computational geometry, it is customary to assume the real-RAM model can be
extended to allow the computation of roots of constant degree polynomials.
We distance ourselves from this practice and take particular care
of using the real-RAM model and the bounded-degree algebraic decision tree
model with only the four arithmetic operators.

