%\chapter{Bird's Eye View}
\chapter{In a Dozen Pages}
\markright{0. In a Dozen Pages}

\todo{Avoid definitions at all cost.}

\todo{Roadmap link to later chapters.}

This thesis is a compilation of the contributions from four papers:
%
\begin{enumerate}
	\item[\ref{paper:ksum-algorithm}] ``\emph{Solving \(k\)-SUM Using Few
		Linear Queries}''
		with Jean Cardinal and John Iacono~\cite{CIO16}.
	\item[\ref{paper:3pol-algorithm}] ``\emph{Subquadratic Algorithms for
		Algebraic 3SUM}''
		with Luis Barba, Jean Cardinal, John Iacono, Stefan Langerman, and Noam Solomon~\cite{BCILOS19}.
	\item[\ref{paper:order-type-encoding}] ``\emph{Subquadratic Encodings for
		Point Configurations}''
		with Jean Cardinal, Timothy Chan, John Iacono, and Stefan Langerman~\cite{CCILO19}.
	\item[\ref{paper:3sum-encoding}] ``\emph{Encoding 3SUM}''
		with Sergio Cabello, Jean Cardinal, John Iacono, Stefan Langerman,
		and Pat Morin~\cite{CCILMO19}.
\end{enumerate}
%
Those contributions have been presented at
the \emph{European Symposium on Algorithms} (ESA),
the \emph{International Symposium on Computational Geometry} (SoCG),
the \emph{Computational Geometry's Young Researchers Forum} (CG:YRF,
a satellite event of SoCG),
the \emph{European Workshop on Computational Geometry} (EuroCG),
and,
the \emph{Annual Fall Workshop on Computational Geometry} (FWCG).

Paper~\ref{paper:ksum-algorithm} was presented at CG:YRF 16 and ESA 16.
%
Paper~\ref{paper:3pol-algorithm} was presented at EuroCG 17 and SoCG 17.
It is published in \emph{Discrete \& Computational Geometry} (DCG).
%
Paper~\ref{paper:order-type-encoding} was presented at FWCG 17, EuroCG 18, and SoCG 18.
It is published in the \emph{Journal of Computational Geometry} (JoCG).
%
Paper~\ref{paper:3sum-encoding} was presented at EuroCG 19.

We begin with an overview of the studied topic.
%
We explain the context in which those papers were written and expose
the contributions contained in each of them.

\section*{Degeneracy Testing Problems}
\input{text/summary/degeneracy-testing-problems}

\section*{GPT}
\input{text/summary/gpt}

\section*{Nonuniform Algorithms}
\input{text/summary/nonuniform-algorithms}

\section*{3SUM}
\input{text/summary/3sum}

\section*{\(k\)-SUM}
\input{text/summary/ksum}
\paragraph{Paper~\ref{paper:ksum-algorithm}}
\input{text/summary/contrib-ksum}

\section*{Sorting}
\input{text/summary/sorting}

\section*{A Zoo of Generalizations}
\input{text/summary/generalization-zoo}

%\section*{SUBSET-SUM}
%Note that the unparameterized version of \(k\)-SUM is the SUBSET-SUM problem,
%which is NP-complete... Discuss nonuniform polytime algo.

%Because of that, we put Sorting and Element Uniqueness as problems at the
%origin of a zoo of problems.

% DONE
% Mention Moshe and Chan's paper at some point and show how it applies to
% other fields (string indexing).}

\section*{Intermediate Problems}
\input{text/summary/intermediate-problems}
\paragraph{Paper~\ref{paper:3pol-algorithm}}
\input{text/summary/contrib-intermediate-problems}

\section*{Encodings}
\input{text/summary/encodings}
\paragraph{Papers~\ref{paper:order-type-encoding}~\&~\ref{paper:3sum-encoding}}
\input{text/summary/contrib-encodings}

\section*{The Secret Ingredient}
\input{text/summary/geometric-divide-and-conquer}

\section*{Wrapping it up}
\input{text/summary/wrap-up}
